Combined effects of protein expression variance and correlation on multicomponent systems

### 2020-03-12 Manuscript/Manuscript.Rmd
.

output:

  pdf_document:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
    includes:
      in_header: header.tex

  html_document: null

  word_document: null
documentclass: article
capsize: normalsize
geometry: margin=1in
fontsize: 11pt
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
spacing: doublespacing
bibliography: 
  - references.bib
  - knitcitations.bib
csl: cell.csl
abstract: Protein expression variation leads to phenotypic variance between cells. This has been demonstrated in cell signaling and differentiation decisions. Additionally coordinated expression of proteins between cells can tune signaling pathways either towards a more binary or analog modality. Though there is some evidence in bulk cell measurements and in bacteria that certain heteromeric subunits or metabolic pathways may be expressed in a coordinated fashion (i.e. operons in bacteria), there has not been a direct measurement of coordinated expression of proteins (independent of TFs) in metazoans (vertebrates?). Here, we measure cell-to-cell variability of relative protein abundance using quantitative proteomics of individual *Xenopus laevis* eggs and show that proteins involved in metabolic pathways or members of heteromeric complexes tend to have high correlations with other members of those pathways/complexes. Our previous work highlighted the fact that correlated expression increases the total variation of a pathway, so one would reason that certain pathways or complexes would need to compensate of this extra source of by reducing the variation of expression of these pathways or complexes. To test this we computed total variance score that took into account both the coefficient of variance and correlations between proteins in a pathway and found that the lower 10% of GO terms were highly enriched for metabolic pathways. When we looked at the relationship between CV and R between these GO terms we found a negative relationship between them, demonstrating that increased correlation needs to come at the expense of decreased variance. Simple molecular models of heteromeric complexes and metabolic pathways demonstrate that this tradeoff can result in higher efficiencies in both function and reduced energy waste. Together, our study argues for a control principle whereby coordinated expression of proteins in a pathway can require lower variance in order to reduce the overall pathway variance, enabling accurate control of active complexes or metabolic pathway activity.
---

@@

## Total variance metric reveals certain modules enforce a trade-off between protein co-expression and noise

The sum of variance for a group of proteins $X_{1},\dotsc,X_{n}$ with dependent relationships:

$$\sigma_{{}_{total}}^2 = \sum_{i=1}^{n} \sigma_{X_{i}}^2 + 2\sum_{i,j:i<j}^{n} \rho(X_{i},X_{j})\sigma_{X_i}\sigma_{X_j}$$

$$\sigma_{{}_{norm}} = \dfrac{\sqrt{\sigma_{{}_{total}}}}{n}$$

When calculating the total variance of a pathway or general group of proteins, the total variance is the square root of the sums of the squares of the coefficient of variance. However, if any of the proteins are correlated in expression, an extra term must be added to account for this correlation. This means that if a group of proteins have a positive correlation, then the total variance will be higher than if there was no correlation. In order to study how variance and coordinated expression are related between single cells, we wanted to calculate a normalized total variance metric. Since we were not interested in the increase of variance due to the number of proteins in a module, we calculated the total variance using the coefficient of variation and correlation coefficients and normalize to the total number of proteins in the module. When total variance is plotted against the number of proteins, there is a clear linear relationship between the two. Once we normalize to the total number of proteins this relationship is lost and we are left with the effects of coefficient of variation and correlation.

@@

In order to see if variation was a regulated process, we grouped the protein variation by gene ontology terms (GO terms) and plotted them in ranked order. We saw that in general, processes xxx were…

To investigate whether coordinated expression goes beyond just protein pairs, we wanted to see if pathways, complexes, and other modules had proteins that with coordinated expression. To answer this question, we grouped proteins by GO terms and calculated the median correlation coefficient across all pairs of proteins within the term. When the ranked media correlation is plotted it’s clear that there are a significant number of GO terms with higher than expected levels of coordinate expression.

>>>>

output:
  pdf_document:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    keep_tex: yes
    number_sections: no
    template: manuscript.latex
    includes:
      in_header: header.tex

  word_document: null
  html_document: null

documentclass: article
capsize: normalsize
geometry: margin=1in
fontsize: 11pt
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
spacing: doublespacing
bibliography: 
  - references.bib
  - knitcitations.bib
csl: cell.csl
abstract: Protein expression variation leads to phenotypic variance between cells. This has been demonstrated in cell signaling and differentiation decisions. Additionally coordinated expression of proteins between cells can tune signaling pathways either towards a more binary or analog modality. Though there is some evidence in bulk cell measurements and in bacteria that certain heteromeric subunits or metabolic pathways may be expressed in a coordinated fashion (i.e. operons in bacteria), there has not been a direct measurement of coordinated expression of proteins (independent of TFs) in metazoans (vertebrates?). Here, we measure cell-to-cell variability of relative protein abundance using quantitative proteomics of individual *Xenopus laevis* eggs and show that proteins involved in metabolic pathways or members of heteromeric complexes tend to have high correlations with other members of those pathways/complexes. Our previous work highlighted the fact that correlated expression increases the total variation of a pathway, so one would reason that certain pathways or complexes would need to compensate of this extra source of by reducing the variation of expression of these pathways or complexes. To test this we computed total variance score that took into account both the coefficient of variance and correlations between proteins in a pathway and found that the lower 10% of GO terms were highly enriched for metabolic pathways. When we looked at the relationship between CV and R between these GO terms we found a negative relationship between them, demonstrating that increased correlation needs to come at the expense of decreased variance. Simple molecular models of heteromeric complexes and metabolic pathways demonstrate that this tradeoff can result in higher efficiencies in both function and reduced energy waste. Together, our study argues for a control principle whereby coordinated expression of proteins in a pathway can require lower variance in order to reduce the overall pathway variance, enabling accurate control of active complexes or metabolic pathway activity.
---

@@

## Variance sum law reveals modules that utilize a trade-off between protein co-expression and noise

Multicomponent modules such as heteromeric protein complexes and metabolic and signaling pathways have a certain tolerance for noise. Two components of the noise of these systems are the expression variance of the individual proteins, as well as the correlation or covariance of expression between the proteins. This means that if a group of proteins have a positive correlation, then the total variance will be higher than if there was no correlation. This single cell proteomics dataset gave us an excellent opportunity to see how variance and co-expression of proteins are related in low variance and high variance modules. The total variance of a module can be calculated using the variance sum law, where for a group of proteins $X_{1},\dotsc,X_{n}$, the total variance of the module is:
$$\sigma_{{}_{total}}^2 = \sum_{i=1}^{n} \sigma_{X_{i}}^2 + 2\sum_{i,j:i<j}^{n} \rho(X_{i},X_{j})\sigma_{X_i}\sigma_{X_j}$$

To categorize the measured protein expression data into modules, we utilized GO term and KEGG pathway annotations. Since pathways and complexes have varying numbers of proteins, and the number of proteins in a module is highly influential on the total variance sum, we normalized the total variance sum by the number of components (Fig. xA and xB). With this normalized total variance metric, we could separate modules into high and low variance categories (Fig 2A). Since co-expression of proteins (positive correlation between protein expression) increases the total variance, we expected to see the high variance modules to be enriched for high levels of both protein expression variance and co-expression (Fig 1B and 1C, red). However, we were suprised to see that the low variance modules were also enriched for high levels of co-expression of proteins, with some of the mean correlation coefficients of modules as high as those in the high variance category (Fig 1B, orange shaded region).

We wondered how these protein modules dealted with the increased variance load caused by the high numbers of correlated proteins, so we looked at the pairwise relationship between the mean variance of the modules and the mean correlations (Fig S1). The low variance modules uniquely showed a significant negative correlation (R = 0.49) between these two metrics, meaning that for modules that enforce a low level of total variance, co-expression of proteins within that module must also have low expression variance.

@@

We saw that in general, processes xxx were… (Fig. 2D).
(Fig. 3A-D) (Fig. 3E-H)
